---
title: "Assignment 1"
author: "Your Name Here"
date: "`r format(Sys.time(), '%d %h, %Y, %I:%M %p')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions:

1. Update your name in the header block, example: `author: "Alex Fout" `
1. Select `File > Save as` and save the file by adding your last name at the beginning with an underscore, example: `fout_assignment_1.rmd`
1. Follow the instructions below to fill in the assignment.
1. Be sure to _run your code chunks when you make them_, to make sure everything works!
1. When you've completed the assignment, __knit__ the document and make sure the resulting HTML or PDF file looks alright.
1. Upload the PDF or HTML file to Canvas (Don't upload the Rmd document).


# Assignment

In this assignment, we'll focus on familiarity with RMarkdown documents and writing R commands.


## Problem 1. 

Here is a code chunk with no code in it.

```{r}

```

Type `1+1` into the above code chunk (don't include the backticks) and press the green play button to the right to run the code.


## Problem 2. 
 
Here's another code chunk:

```{r}
12 * 2
```

Change the code so that it's computing 12 _divided by_ 2, and press the green play button to run the code.


## Problem 3. 

Type code in the code chunk below to find the result of 22 divided by 7.

```{r}

```

Add a comment inside the code chunk _before_ the command you just wrote, and use the comment to explain what your code is doing.


## Problem 4.

Create a new code chunk below and compute the result of 12 squared.
Make sure that code chunk works by running the code chunk.


## Problem 5.

Create another code chunk below and write code to print out the message: Hello from the other side!


## Problem 6.

Here is an example of a hyperlink in RMarkdown:

[The R Graph Gallery](https://www.r-graph-gallery.com/)

The part in square brackets, [], is the text that will be displayed after knitting.
The part in parentheses, (), is the URL to a website.

VIsit the R Graph Gallery by following the link above, and spend some time perusing the different categories. 
You can click on a graph type (for example, "violin") and scroll down to view some examples of that type. 

Find three different examples graphs from three different categories that you find particularly interesting.
In the sections below, create a hyperlink which links to the example graph you found.


### Example 1


### Example 2


### Example 3



## Problem 7. 

R is used in many different disciplines!
Take your major (sociology for example) and try doing the following google search:

"R for sociology" (or whatever your major is)

Look at a few search results to see if there are some interesting resources available to people in your field who are using R.
If you aren't finding anything very promising, try something like:

"sociology with R"

Pick a website that seems interesting to you, and create a hyperlink below for it.
If you can't find any _good_ examples, repeat the search with a different example from yours, and see if you find something interesting. 

### Hyperlink


## Problem 8.

Visit [#rstats](https://twitter.com/hashtag/rstats?lang=en) on twitter and peruse the posts for a while.
Find one post that looks interesting to you and link to it below:

### Hyperlink


## End

This is the end of the assignment!
You can knit the document and upload it to Canvas


