---
title: "Assignment 1"
author: "Your Name Here"
date: "`r format(Sys.time(), '%d %h, %Y, %I:%M %p')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions:

1. Update your name in the header block, example: `author: "Alex Fout" `
1. Select `File > Save as` and save the file by adding your last name at the beginning with an underscore, example: `fout_assignment_1.rmd`
1. Follow the instructions below to fill in the assignment.
1. Be sure to _run your code chunks when you make them_, to make sure everything works!
1. When you've completed the assignment, __knit__ the document and make sure the resulting HTML or PDF file looks alright.
1. Upload the PDF or HTML file to Canvas (Don't upload the Rmd document).


# Assignment

In this assignment, we'll focus on familiarity with RMarkdown documents and writing R commands.

## 1. 

Here is a code chunk with no code in it.

```{r}

```

Type `1+1` into the above code chunk (don't include the backticks) and press the green play button to the right to run the code.

## 2. 

Here's another code chunk:

```{r}

```
## 3. 

Enter code in the code chunk to find the result of 22 divided by 7

```{r}

```

## 4.

Create a new code chunk below and compute the result of 12 squared.
Make sure that code chunk works by running the code chunk.



## 5.

Create another code chunk below and write code to print out the message: Hello from the other side!



## 6.

Here is an example of a hyperlink in RMarkdown:

[The R Graph Gallery](https://www.r-graph-gallery.com/)

VIsit the R Graph Gallery by following the link above, and 


