---
title: "Progress Check 2"
author: "Your Name Here"
date: "`r format(Sys.time(), '%d %h, %Y, %I:%M %p')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions:

1. Update your name in the header block, example: `author: "Alex Fout" `
1. Select `File > Save as` and save the file by adding your last name at the beginning with an underscore, example: `fout_progress_check_2.rmd`
1. Each time you encounter a purple "Progress Check" block in the textbook, create a new sub-subheading  and a new code block for any code you write.
1. Be sure to _run your code chunks when you make them_, to make sure everything works!
1. Your completed assignment should contain all progress checks from Chapter 3 through (and including) Section 4.3
1. When you've completed the assignment, __knit__ the document and make sure the resulting HTML or PDF file looks alright.
1. Upload the PDF or HTML file to Canvas (Don't upload the Rmd document).


# Example:

If you're in section 3.1, create a section 3.1 subheading with two #'s:

## Section 3.1

Each time you encounter a progress check block in this section, make a sub-subheading with three #'s, and __use the number in the lower right corner of the block__

### 1

For each progress check, you'll create a code chunk for any code you write:

```{r}
print("here's some code for the first progress check box in 3.1")
```

### 2

```{r}
print("here's some code for the second progress check box in 3.1")
```

## Section 3.2

Here's where section 3.2 progress checks will go, etc.


# Progress Checks:

You fill in this part!
