---
title: "Progress Check 2"
author: "Your Name Here"
date: "`r format(Sys.time(), '%d %h, %Y, %I:%M %p')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions:

1. Update your name in the header block, example: "Alex Fout" 
1. Each time you encounter a purple "Progress Check" block in the textbook, create a new code block with the same title 
1. Be sure to run your code chunks when you make them, to make sure everything works!
1. Your completed assignment should contain all progress checks from section XX through section XX
1. When you've completed the assignment, **knit** the document and upload to Canvas.


## Example:

If you're in section 3.1, create a section 3.1 subheading with two #'s:

## Section 3.1

Each time you encounter a progress check block in this section, make a sub-subheading with three #'s, and __use the number in the lower right corner of the block__

### 1

For each progress check, you'll create a code chunk for any code you write:

```{r}
print("here's some code")
```

### 2

This will


## Progress Checks:

You fill in this part!
