---
title: "Progress Check 3"
author: "Your Name Here"
date: "`r format(Sys.time(), '%d %h, %Y, %I:%M %p')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions:

1. Update your name in the header block, example: `author: "Alex Fout" `
2. Select `File > Save as` and save the file by adding your last name at the beginning with an underscore, example: `fout_progress_check_3.rmd`
3. Each time you encounter a purple "Progress Check" block in the textbook, create a new sub-subheading  and a new code block for any code you write.
4. Be sure to _run your code chunks when you make them_, to make sure everything works!
5. Your completed assignment should contain all progress checks from section 4.4 through and including section 5.4
6. When you've completed the assignment, __knit__ the document and make sure the resulting HTML or PDF file looks alright.
7. Upload the PDF or HTML file to Canvas (Don't upload the Rmd document).


# Example:

If you're in section 3.1, create a section 3.1 subheading with two #'s:

## Section 3.1

Each time you encounter a progress check block in this section, make a sub-subheading with three #'s, and __use the number in the lower right corner of the block__

### 1

For each progress check, you'll create a code chunk for any code you write:

```{r}
2 + 2
```

And you may also include some extra text to answer any questions not requiring code.

### 2

```{r}
print("here's some code for the second progress check box in 3.1")
```

## Section 3.2

Here's where section 3.2 progress checks will go, etc.

# R Markdown Tips:

- Save your work often!
- Knit your work often!
- Make sure any headings have a blank line _before_ them. E.g. the following will not work:

this text is too close to the heading
## this is a heading

To see the problem, knit the document and observe that "this is a heading" is not displayed as a heading.


# Progress Checks:

You fill in this part!