# Working with Data

Congratulations! You're all set up and ready to work with data. That's what we're all here for, right? 

There's a lot to take in while you're getting started, so don't worry if you need to look up the command names or documentation. 

It's important to make good habits when you're starting out, but keep in mind that R is constantly being updated and new packages come out frequently. There are also many ways to get to the same place sometimes, so feel free to get your feet dirty experimenting! 

```{block, type="caution"}
This section is still under construction
```
## Base R and the Tidyverse

So far, you've been working with commands that are included when you download R, called [base R](https://rstudio.com/wp-content/uploads/2016/05/base-r.pdf). There are many packages that can go above and beyond base R (an advantage of open-source code), and we'll start by installing a core set of them, called _*the tidyverse*_. The packages included in [the tidyverse](https://www.tidyverse.org/) are ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, and forcats. Each of these serves a different purpose, and you'll get more familiar with them as you begin to write code. 

>The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.

To get analyze data, you will often need to wrangle it from the raw srtructure into your desired format. Let's get wrangling!

<iframe src="https://giphy.com/embed/g0QGZxpbFueYVluQb4" width="480" height="368" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/georgestrait-george-strait-heartland-g0QGZxpbFueYVluQb4">via GIPHY</a></p>

## Loading / Saving Data

```{block, type="caution"}
This section is still under construction
```


### "Taster"(?) list of file forms and sources

### reading/writing csv

_*Should we mention using the UI or stick to the code?*_ -liz

Let's import some data about animal crossing from the tidytuesday github repo. We'll use the package readr here. The read_csv() function from the readr package is is an example of a command that is faster/better than the similar read.csv() function in base R.

```{r, message = FALSE}
library(tidyverse) # this loads all of the packages in the tidyverse
villagers <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-05/villagers.csv') 
# `readr::` tells us this command is from the readr package. It's not necessary if there is no conflicting command (ex. another package has a command called `read_csv()` but here it gives us context. Using read_csv() will work just as well.
```

Okay, so we have an object named `villagers` now, but how do we see what's inside? 

(Would you rather have this is this section or "Practice" or as an assignment? -liz)

```{r}
str(villagers)
```

### Best practices

principles of Tidy Data (wickham 2014)

raw data as read only 

```{block, type="feedback"}
Give me some [feedback](https://forms.gle/2Fw8y4aY51ijdmyE9) about the content in this section.
```

## Downloading and Saving 

This section seems like the same as Loading / Saving Data? -liz

```{block, type="caution"}
This section is still under construction
```
```{block, type="video"}
downloading and saving example csv from canvas or website
```
```{block, type="feedback"}
Give me some [feedback](https://forms.gle/2Fw8y4aY51ijdmyE9) about the content in this section.
```

## Working With Data

```{block, type="caution"}
This section is still under construction
```

### Basic indexing

matrices: can use single index

nested indexing: [[1]][3]

### Advanced indexing

```{block, type="bonus"}
dplyr example
```
```{block, type="feedback"}
Give me some [feedback](https://forms.gle/2Fw8y4aY51ijdmyE9) about the content in this section.
```

### Summarizing vectors

Mean, std, etc.

### Summarizing matrices

### Summarizing vectors

### Basic Plotting

```{block, type="bonus"}
ggplot example
```
```{block, type="caution"}
Remember to pay close attention to the syntax when using ggplot. R won't work if parenthesis or quotes are not paired. If you're adding multiple layers, put each on a new line.

Adding structure to make your code more readable is like doing your future self a favor (and anyone reading your code).
```

## Practice

```{block, type="caution"}
This section is still under construction
```

```{block, type="progress"}
asdf
```

