# Working with Data

```{block, type="caution"}
This chapter is still under construction
```

In this Chapter, we'll use R to do interesting things with _data_.
This is by far the most popular use of the R programming language, and arguably the most fun!
You'll learn how to load data sets into R, do interesting things with them, and save your results.

## Quick Example

Before diving into detail, let's do a quick example so you can begin to see what is possible with data in R.
As we mentioned in the last chapter, R includes some pre-packaged data sets, which you can access with the `data()` command.
One of the data sets is `Seatbelts`, which documents road casualties in Great Britain between 1969 and 1984.
Firstly, we need to convert `Seatbelts` to a dataframe, because it starts out as a "Time-Series", which we haven't discussed.
```{r}
Seatbelts <- data.frame(as.matrix(Seatbelts), date=time(Seatbelts))
```
We've also added a month and year column 

look at the dimensions of the data set with the `dim` command:
```{r}
dim(Seatbelts)   # get the number of dimensions in the Seatbelts dataframe
```
This shows that there are 192 rows (months), and 9 columns (variables measured each month).
We could also determine the number of rows and columns separately using the `nrow` and `ncol` functions.
To view the first few rows of the Seatbelts dataframe, use the `head` command:

```{r}
head(Seatbelts)  # view first few rows of the Seatbelts dataset
```

This is a good way to learn which variables are being measured (columns) and see some example observations (rows) for each variable.
Because these data are included with R, more information about each variable can be found with:

```{r, eval=F}
?Seatbelts
```

Next, let's view a summary of each column with the summary function:

```{r}
summary(Seatbelts)
```

Since each column is numeric, R shows a five number summary (minimum, first quartile, median, third quartile, maximum) and mean for each column. 
We learn, for example, that the average number of drivers killed per month is 1670, and that the greatest price of petrol was £0.13 per litre!
Let's view a histogram of `DriversKilled`:

```{r, fig.cap="Histogram of Drivers Killed in Seatbelt data"}
hist(Seatbelts$DriversKilled, breaks=20)
```
We see that in some months, more than 150 drivers were killed!
We can calculate how many exactly like so:

```{r}
sum(Seatbelts$DriversKilled > 150)
```
To investigate the effect of the seat belt law, let's create a scatter plot Drivers killed against time:
```{r, fig.keep='all', fig.cap="UK Seatbelt deaths vs time"}
plot(Seatbelts$date, Seatbelts$DriversKilled)
```
By adding a `col` argument to the `plot` function, we can color the points based on whether the law was in effect:

```{r, fig.keep='all', fig.cap="UK Seatbelt deaths vs time, red = no seatbelt law, green = seatbelt law"}
plot(Seatbelts$date, Seatbelts$DriversKilled, col=(Seatbelts$law+2))
```

There do appear to be fewer deaths, but there is so much fluctuation in deaths each year that it's difficult to tell.
Let's change the x-axis to reflect month of the year instead of date:

```{r, fig.cap="UK Driver Deaths vs. Month"}
plot((Seatbelts$date %% 1) * 12 + 1, Seatbelts$DriversKilled, 
     xlab = "Month", col=Seatbelts$law + 2)
```
This plot shows that there is a clear seasonal effect in the number of deaths with higher deaths occurring in the Fall/Winter compared to Spring/Summer.
We can also see that within each month, the traffic deaths after enacting the Seatbelt law are among the lowest.

```{block, type="progress"}
Another data set included in R is `mtcars`. Following the example above, find the dimension of `mtcars` and have R print out a summary of each column, then create a scatter plot of fuel economy (`mpg`) to engine displacement.
What do you observe about the relationship between these two variables?
```

This concludes the quick example.
In the rest of this chapter, we'll talk in more detail about the different steps of working with data, and how to complete them using R!

```{block, type="reflect"}
People often use data in order to answer questions, but often times, learning about data can generate even _more_ questions than it answers.
Take a moment to think of a question that you have about the `Seatbelts` dataset. 
Do you think the question can be answered using the data alone?
If not, what other sources of data might be available which can help answer the question?
```

```{block, type='feedback'}
Any feedback for this section? Click [here](https://docs.google.com/forms/d/e/1FAIpQLSePQZ3lIaCIPo9J2owXImHZ_9wBEgTo21A0s-A1ty28u4yfvw/viewform?entry.1684471501=Quick%20Example)
``` 

## Loading / Saving Data


```{block, type="caution"}
This section is still under construction
```

### "Taster"(?) list of file forms and sources

### Data Format

Talk about what columns and rows mean 

(long vs. wide? probably not...)

### reading/writing csv

### Best practices

principles of Tidy Data (wickham 2014)

raw data as read only

```{block, type="feedback"}
Give me some [feedback](https://forms.gle/2Fw8y4aY51ijdmyE9) about the content in this section.
```

## Downloading and Saving

```{block, type="caution"}
This section is still under construction
```
```{block, type="video"}
downloading and saving example csv from canvas or website
```
```{block, type="feedback"}
Give me some [feedback](https://forms.gle/2Fw8y4aY51ijdmyE9) about the content in this section.
```

## Working With Data

```{block, type="caution"}
This section is still under construction
```

### Basic indexing

matrices: can use single index

nested indexing: [[1]][3]

### Advanced indexing

```{block, type="bonus"}
dplyr example
```
```{block, type="feedback"}
Give me some [feedback](https://forms.gle/2Fw8y4aY51ijdmyE9) about the content in this section.
```

### Summarizing vectors

Mean, std, etc.

### Summarizing matrices

### Summarizing vectors

### Basic Plotting

```{block, type="bonus"}
ggplot example
```
```{block, type="caution"}
Remember to pay close attention to the syntax when using ggplot. R won't work if parenthesis or quotes are not paired. If you're adding multiple layers, put each on a new line.

Adding structure to make your code more readable is like doing your future self a favor (and anyone reading your code).
```

## Practice

```{block, type="caution"}
This section is still under construction
```

```{block, type="progress"}
asdf
```

